Open archaeology, open source?

### 2023-10-11 analysis/_04-open_archaeology.qmd
Merge conflicts

```{=html}
<!-- Results I: description of open archaeology projects

- What kind of projects comprise open archaeology?
  x Breakdown of tags and categories
  x Breakdown of platforms and languages
  x Tags by category
  x Licenses?
- How has this changed over time?
  - Growth of open archaeology by number of commits
  - Growth of open archaeology by number of repos
  - Compare changing proportions of each category on a year by year basis
  - Identify periods associated with different rates of growth, and relate these with different general attitudes in the history of digital archaeology
-->
```
As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
This primarily constitutes software but also includes various forms of open documents. @tbl-categories summarizes the kinds if resources that appear in open-archaeo, and breaks them down into more precise categories.

@@

As per @tbl-platforms, The statistical programming language R `r percent(p_platform_r)` is overwhelmingly the most common platform among projects that extend upon existing programming languages and applications.
This is followed by Python <!-- percent -->, which is another popular scientific scripting language, and the open source geographic information system QGIS <!-- percent -->.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

```{r fig-tags}
#| fig-cap: Frequency of tags applied to open archaeology projects included in open-archaeo, broken down by category.
detail_tags <- c("Instrumental Neutron activation analysis",
                 "Harris Matrix",
                 "aDNA Simulators",
                 "Damage profilers",
                 "Metagenomic classifers",
                 "aDNA simulators",
                 "aDNA damage-aware genotypers",
                 "Low-coverage contamination estimation",
                 "Low-coverage adapted population-genetics tools",
                 "aDNA preprocessing",
                 "aDNA aligner",
                 "aDNA assembler",
                 "aDNA toolkit")

oarch %>%
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) %>%
  drop_na(tag) %>%
  filter(!tag %in% detail_tags) %>%
  mutate(category = recode(category,
                                  "Specifications etc." = "Documents",
                                  "Products" = "Documents",
                                  "Guides" = "Documents",
                                  "Lists and datasets" = "Documents",
                                  .default = category)) %>%
  group_by(tag, category) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(total = sum(n)) %>%
  ggplot(aes(fct_reorder(tag, total), n, fill = category)) +
  # facet_wrap(vars(category)) +
  geom_col() +
  scale_fill_bright() +
  coord_flip() +
  labs(
    title = "Tags by category",
    caption = "Data: open-archaeo.info",
    x = NULL, y = "Number of projects",
    fill = NULL
  ) +
  theme_oarch() +
  theme(
    axis.text.y = element_text(size = 7)
  )
```

>>>>

<!-- Results I: description of open archaeology projects

- What kind of projects comprise open archaeology?
  x Breakdown of tags and categories
  x Breakdown of platforms and languages
  x Tags by category
  x Licenses?
- How has this changed over time?
  - Growth of open archaeology by number of commits
  - Growth of open archaeology by number of repos
  - Compare changing proportions of each category on a year by year basis
  - Identify periods associated with different rates of growth, and relate these with different general attitudes in the history of digital archaeology
-->
```

As of writing, open-archaeo catalogues `r nrow(oarch)` resources created by and for archaeologists.
This primarily constitutes software but also includes various forms of open documents. @tbl-categories summarizes the kinds if resources that appear in open-archaeo, and breaks them down into more precise categories.

@@

`r percent(p_platform_r)` of projects are extensions to the statistical programming language R, making it the most widely-used 'platform' by a large margin (@tbl-platforms).
Python, another programming language, is also relatively popular <!-- TODO: percent -->, as are plugins for the open source geographic information system QGIS <!-- TODO: percent -->.
Beyond that, there is a rather fragmented landscape of plugins for other desktop software (e.g. AutoCAD, ArcGIS), a number of lesser used programming languages, and a genre consisting of custom forms and spreadsheet templates.
Many of these are targeted by only one or two developers; the larger platforms tend to be more diverse.

@@

```{r fig-tags}
#| fig-cap: Tags by category
oarch %>%
  pivot_longer(c(tag1, tag2, tag3, tag4, tag5), values_to = "tag",
               values_drop_na = TRUE) %>%
  drop_na(tag) %>%
  mutate(category = recode(category,
                                  "Specifications, protocols and schemas" = "Documents",
                                  "Products" = "Documents",
                                  "Guides" = "Documents",
                                  "Lists and datasets" = "Documents",
                                  .default = category)) %>%
  group_by(tag, category) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(total = sum(n)) %>%
  ggplot(aes(fct_reorder(tag, total), n, fill = category)) +
  geom_col() +
  scale_fill_bright() +
  coord_flip() +
  labs(
    x = NULL, y = "Number of projects",
    fill = NULL
  ) +
  theme_oarch() +
  theme(
    axis.text.y = element_text(size = 7)
  )
```
